Mechanisation of Session-based Lambda Calculus Type Systems

### 2014-11-26 proposal/proposal.tex
Finished read-through and update sections: just pictures left.

The aim of this project is to develop a framework for formalising lambda
calculus type systems involving session types using interactive theorem
proving (in particular, the system Coq). The motivation for the framework is
that it will enable a whole class of related languages and type systems to be
formalised with little extra proof effort. Inspired by previous work on a
mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}, I
shift to the lambda calculus to provide a basis on which session types for
mainstream programming languages can be studied, whilst taking advantage of
previous proof effort and specialising only for those parts unique to the
particular system. The framework will be heavily based on calculi developed by
\citeauthor{Wadler:2012}~\cite{Wadler:2012} describing a relationship between
classical linear logic (CP) and a functional language with session types
(GV). GV is a minimal functional language supporting linearity so I believe
such a basis will be general enough to support reasoning about a large class
of functional languages with similar base properties.

@@

A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and
\citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a
$\lambda$-calculus based language with session types for channel
primitives. The type system is nonstandard, featuring an additional
environment for tracking the session types of channels; these types are
modified during term reduction to reflect communication. The extra channel
environment is provided to support possible aliasing of channels by treating
channel variables as pointers or references into the channel environment. So,
for a channel $c$ of session type $S$, a variable $x$ referring to this
channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the
channel enviornment. The typing judgements and term reduction relation must be
augmented to handle these environments. Unfortunately, a side-effect of this
is that function types must specify the pre and post-states of channel
environments and therefore, the type of arguments cannot work up to
alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$
have the same session type, one cannot define a function to operate on
variables of either channel type since the function definition will explicitly
mention the identifier for the channel (in this case either, $Chan~c$ or
$Chan~d$). However, the system does provide an attractive non-linear channel
type which has been neglected in favour of a linear treatment of channels in
more recent
works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2012}. Certainly, it is the
case that an aliased approach would fit better with mainstream programming
languages and perhaps studying a similar system could offer some insights.

@@

Such a rule would appear to only benefit recursive session types. Indeed,
focussing on a non-recursive channel would be pointless since one would be
prevented from communicating across it since the focus rule requires
maintenance of the channel type. Further, it would be undesirable to permit
modification of a session type without guaranteeing exclusive access to the
channel since this could lead to invalid communication e.g. trying to
communicate on a closed channel.

The Alms language~\cite{Aldrich:2009} provides a functional language with
affine types based on linear logic. A notion of type \textit{kind} is used to
distinguish between affine and unrestricted values. The typing judgements use
two contexts; one for variables with unrestricted kind and another for
variables of affine type and type variables. Functions are also labelled with
a kind which may depend on their arguments e.g. a multiple argument function
must have affine kinding if it accepts an affine typed resource as an argument
other than the final argument. The reason for this is simply to prevent the
function from using an affine resource more than once. Additionally, functions
act as closures over their environments so their qualifier may be determined
by a free variable in the environment, and a relation is provided to calculate
the appropriate kind qualifier. Type variables are provided to support kind
polymorphism of function qualifiers. The system is expressive enough to
implement session types as a module using a similar approach to
\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as
affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms'
type system can support the ``adoption and focus'' system but do not specify
details or whether any changes are necessary to their typing
judgements. Moreover, \citeauthor{Fahndrich:2002} do not provide operational
semantics or soundness results for the ``adoption and focus'' type system.

@@

\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} brings together a
process calculus, CP, based on the earlier work by
\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
\citeauthor{Gay:2010:LAST}. The work describes a continuation-passing style
(CPS) translation from GV to CP treating session types as propositions in a
classical linear logic. The process calculus is slightly different to that
described by $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in
nature, the constructs for server replication are modified; CP provides a
weakening and contraction rules whereas these are encoded in one rule for
process reduction by \citeauthor{Caires:2010:STI}. Additionally, rather than a
two-sided sequent calculus, \citeauthor{Wadler:2012} uses one-sided sequents
for defining the process calculus, which provides a more intuitive
presentation of session type duality and clearer correspondence to a
$\lambda$-calculi surface syntax. As presented, GV does not support all
features of CP and thus the translation is one way (GV to
CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to
provide support for polymorphism and replication, following a similar CPS
translation scheme. Lastly,
\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
semantics for GV and their system has greater similarities with LAST than
prior GV works, for example, by treating send and receive as linear functions
rather than language primitives.

@@

I intend to mechanise the calculi described by
\citeauthor{Wadler:2012}~\cite{Wadler:2012} using the Coq proof assistant. In
the spirit of the POPLMARK challenge, I aim to utilise available libraries and
infrastructure as much as
possible. \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a
metatheory library for handling, among other things, association lists
(environments) and free variable representation. I intend to use the version
distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which has been
updated for Coq version 8.4pl4. \citeauthor{Park:2014:MMW} describe a
technique for removing nonlinear contexts from typing judgements. While their
work extended to \fpop it is not clear how to handle an environment containing
both non-linear and linear types as in GV. I wish to maintain as close a
relationship as possible to the paper system presented by
\citeauthor{Wadler:2012}, so separating out the non-linear and linear
components (as in \fpop) is not an option at this stage.

Following \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, I use the
locally nameless first-order representation for binders and free variables,
preventing variable capture issues present in other first-order
representations. Likewise, I make some simplifying assumptions about variable
substitution, namely that \textit{closed} terms (terms with no unbound
indices) are substituted for variables, which avoids having to define lifting
constructs on the substitute term. The typing rules will not contain a rule
for an unbound index so this decision is sound. A consequence is that we must
define a notion of opening language terms involving binders as described
by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
$t$ which is well-typed if $t$ is well-typed when its unbound index,
representing the argument to the abstraction, is replaced by some fresh
variable $x$, written $t^x$. Some terms in GV require multiple binders but I
believe it will be straightforward to extend this notion. Unfortunately, I
must define similar operations on CP terms and handle substitution and opening
on them as well as propositions which can accept propositional variables. I
hope to be able to generalise as much as possible with the help of the
Metatheory library. Alternatively, I could have chosen a higher-order abstract
syntax (HOAS) approach such as the parametric version (PHOAS) described
by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
on the metalanguage (Coq, in this case) for handling binders and substitution
so has less overhead in managing syntax than the locally nameless
approach. However, variables are now parameters defined at the level of the
metalanguage so can represent terms not valid in the language being
mechanised. Further, a higher-order approach would prevent the use of the
Metatheory library which provides a number of useful tactics for manipulating
typing contexts.

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\cite{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, which
excludes a finite set of variables from being considered as the binder. In
contrast, the traditional ``exists-fresh'' approach, where the binder is only
required to be fresh within the abstraction's body, does not produce strong
enough induction principles in some cases. This change will necessitate
changes to the typing rules of GV and CP. Consider for instance the connect
with rule in GV~\cite{Wadler:2012}:

@@

The conclusion binds a variable

Consider also the rule for output in CP~\cite{Wadler:2012} and its
corresponding version used here:

@@

The rule here ... is this needed? What are we showing here with the different
version?

While the aim is to be faithful to \citeauthor{Wadler:2012}'s description, I
may need to alter the approach slightly to avoid tricky formulations. For
instance, instead of providing n-ary branch and choice in GV, we provide only
binary versions of these operators to simplify the development. This change is
not restrictive however, since CP itself provides only binary versions of plus
($\oplus$) and with ($\with$) constructs. Likewise, on a more technical note,
we amalgamate all types into one inductive definition and provide a predicate
for restricting typing rules to consider only valid session
types. \citeauthor{Wadler:2012} defines GV types as a set of mutually
recursive definitions; session types are types which may contain types as
subcomponents (arguments to send, for example). Unfortunately, handling
mutually inductive definitions in Coq can be quite involved; requiring one to
either rely on the Coq system to provide a stronger mutual induction principle
or defining one manually. Such a definition is possible, but it complicates
elimination via the induction principle.
% ( since the standard induction tactic will not work?)

@@

After December, it remains to define the translation between GV and CP which I
envisage will be the bulk of the work, requiring several rules on typing
derivations; I envisage being able to define these as inductive datatype
constructors but I shall take care to reuse any suitable libraries that are
available, e.g. a library for handling CPS translations.

As discussed previously, it would be
interesting to study a system for supporting side-effecting functions (aka
closures) for session channels. In order to achieve this, operational
semantics for GV would need to be defined, and the recent draft by
\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
do this.

For this step of the work, I allocated one month. For the remainder, I shall
look to extend the definitions of GV and CP using the definitions I have
provided

I am going to look into adapting adoption and focus for the GV setting to
allow one to control aliasing on session typed channels independent of their
session characteristics so the aim is not to manipulate a linear type but to
provide a nonlinear view of some resource and when changes are to be made,
enforce some non-aliased view such that there no references in the context
perhaps the notion I'm looking for is to prevent sending on a guarded type or
enforce the guarded type must be preserved in the adoption block the trick in
the adoption and focus system is that the type will be deallocated when its
adopter's lifetime ends; I suspect in our system this is not useful since we
will only be permitted to use a channel in a non-aliased context

>>>>

The aim of this project is to develop a framework for formalising lambda
calculus type systems involving session types using interactive theorem
proving (in particular, the system Coq). The motivation for the framework is
that it will enable a whole class of related languages and type systems to be
formalised with little extra proof effort. Inspired by previous work on a
mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}, I
shift to the lambda calculus to provide a basis on which session types for
mainstream programming languages can be studied, whilst taking advantage of
previous proof effort and specialising only for those parts unique to the
particular system. The framework will be heavily based on calculi developed by
\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
classical linear logic (CP) and a functional language with session types
(GV). GV is a minimal functional language supporting linearity so I believe
such a basis will be general enough to support reasoning about a large class
of functional languages with similar base properties.

@@

A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and
\citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a
$\lambda$-calculus based language with session types for channel
primitives. The type system is nonstandard, featuring an additional
environment for tracking the session types of channels; these types are
modified during term reduction to reflect communication. The extra channel
environment is provided to support possible aliasing of channels by treating
channel variables as pointers or references into the channel environment. So,
for a channel $c$ of session type $S$, a variable $x$ referring to this
channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the
channel enviornment. The typing judgements and term reduction relation must be
augmented to handle these environments. Unfortunately, a side-effect of this
is that function types must specify the pre and post-states of channel
environments and therefore, the type of arguments cannot work up to
alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$
have the same session type, one cannot define a function to operate on
variables of either channel type since the function definition will explicitly
mention the identifier for the channel (in this case either, $Chan~c$ or
$Chan~d$). However, the system does provide an attractive non-linear channel
type which has been neglected in favour of a linear treatment of channels in
more recent
works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2014}. Certainly, it is the
case that an aliased approach would fit better with mainstream programming
languages and perhaps studying a similar system could offer some insights.

@@

Such a rule would appear to only benefit recursive session types. Indeed,
focussing on a non-recursive channel would be pointless since one would be
prevented from communicating across it since the focus rule requires
maintenance of the channel type. Further, it would be undesirable to permit
modification of a session type without guaranteeing exclusive access to the
channel since this could lead to invalid communication e.g. trying to
communicate on a closed channel. Additionally, a linear object is consumed
upon being adopted and will be deallocated when its adopter's lifetime ends,
although an alternative semantics could be defined which returned the linear
resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
operational semantics or soundness results for the ``adoption and focus'' type
system.

The Alms language~\cite{Aldrich:2009} provides a functional language with
affine types based on linear logic. A notion of type \textit{kind} is used to
distinguish between affine and unrestricted values. The typing judgements use
two contexts; one for variables with unrestricted kind and another for
variables of affine type and type variables. Functions are also labelled with
a kind which may depend on their arguments e.g. a multiple argument function
must have affine kinding if it accepts an affine typed resource as an argument
other than the final argument. The reason for this is simply to prevent the
function from using an affine resource more than once. Additionally, functions
act as closures over their environments so their qualifier may be determined
by a free variable in the environment, and a relation is provided to calculate
the appropriate kind qualifier. Type variables are provided to support kind
polymorphism of function qualifiers. The system is expressive enough to
implement session types as a module using a similar approach to
\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as
affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms'
type system can support the ``adoption and focus'' system but do not specify
details or whether any changes are necessary to their typing judgements.

@@

\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
process calculus, CP, based on the earlier work by
\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
\citeauthor{Gay:2010:LAST}. The work describes a continuation-passing style
(CPS) translation from GV to CP treating session types as propositions in a
classical linear logic. The process calculus is slightly different to that
described by $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in
nature, the constructs for server replication are modified; CP provides a
weakening and contraction rules whereas these are encoded in one rule for
process reduction by \citeauthor{Caires:2010:STI}. Additionally, rather than a
two-sided sequent calculus, \citeauthor{Wadler:2014} uses one-sided sequents
for defining the process calculus, which provides a more intuitive
presentation of session type duality and clearer correspondence to a
$\lambda$-calculi surface syntax. As presented, GV does not support all
features of CP and thus the translation is one way (GV to
CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to
provide support for polymorphism and replication, following a similar CPS
translation scheme. Lastly,
\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
semantics for GV and their system has greater similarities with LAST than
prior GV works, for example, by treating send and receive as linear functions
rather than language primitives.

@@

I intend to mechanise the calculi described by
\citeauthor{Wadler:2014}~\cite{Wadler:2014} using the Coq proof assistant. In
the spirit of the POPLMARK challenge, I aim to utilise available libraries and
infrastructure as much as
possible. \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a
metatheory library for handling, among other things, association lists
(environments) and free variable representation. I intend to use the version
distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which has been
updated for Coq version 8.4pl4. \citeauthor{Park:2014:MMW} describe a
technique for removing nonlinear contexts from typing judgements. While their
work extended to \fpop it is not clear how to handle an environment containing
both non-linear and linear types as in GV. I wish to maintain as close a
relationship as possible to the paper system presented by
\citeauthor{Wadler:2014}, so separating out the non-linear and linear
components (as in \fpop) is not an option at this stage.

Following \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, I use the
locally nameless first-order representation for binders and free variables,
preventing variable capture issues present in other first-order
representations. Likewise, I make some simplifying assumptions about variable
substitution, namely that \textit{closed} terms (terms with no unbound
indices) are substituted for variables, which avoids having to define lifting
constructs on the substitute term. The typing rules will not contain a rule
for an unbound index so this decision is sound. A consequence is that we must
define a notion of opening language terms involving binders as described
by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
$t$ which is well-typed if $t$ is well-typed when its unbound index,
representing the argument to the abstraction, is replaced by some fresh
variable $x$, written $t^x$. Some terms in GV require multiple binders but I
believe it will be straightforward to extend this notion. Unfortunately, I
must define similar operations on CP terms and handle substitution and opening
on them as well as propositions which can accept propositional variables. I
hope to be able to generalise as much as possible with the help of the
Metatheory library.

Alternatively, I could have chosen a higher-order abstract syntax (HOAS)
approach such as the parametric version (PHOAS) described
by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
on the metalanguage (Coq, in this case) for handling binders and substitution
so has less overhead in managing syntax than the locally nameless
approach. However, variables are now parameters defined at the level of the
metalanguage so can represent terms not valid in the language being
mechanised. Further, a higher-order approach would prevent the use of the
Metatheory library which provides a number of useful tactics for manipulating
typing contexts.

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\cite{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, which
excludes a finite set of variables from being considered as the binder. In
contrast, the traditional ``exists-fresh'' approach, where the binder is only
required to be fresh within the abstraction's body, does not produce strong
enough induction principles in some cases. This change will necessitate
changes to the typing rules of GV and CP. Consider for instance the Connect
rule in GV~\cite{Wadler:2014} (Connect) and its corresponding cofinite
quantification version (ConnectCO):

@@

The conclusion binds a variable in Connect but a type in ConnectCO. Thus, in
the premise for ConnectCO one can choose a suitably fresh $x$. The same change is made to the rules for CP so that, for example, output which allocates a fresh variable $y$ no longer names $y$ in the conclusion:

@@

Since the name $y$ is assumed to be fresh in the original presentation these
changes do not alter the typing of processes nor the semantics of the
translation from GV to CP. An equivalence between ``exists-fresh'' and
cofinite quantification definitions of the simply-typed $\lambda$ calculus is
given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. In general, I
do not aim to formally prove adequacy of the mechanisation.

While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
may need to alter the approach slightly to avoid tricky formulations. For
instance, instead of providing $n$-ary branch and choice in GV, we provide
only binary versions of these operators to simplify the development. This
change is not restrictive however, since CP itself provides only binary
versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
technical note, I amalgamate all types into one inductive definition and
provide a predicate for restricting typing rules to consider only valid
session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
recursive definitions; session types are types which may contain types as
subcomponents (arguments to send, for example). Unfortunately, handling
mutually inductive definitions in Coq can be quite involved; requiring one to
either rely on the Coq system to provide a stronger mutual induction principle
or defining one manually. Such a definition is possible, but it complicates
elimination via the induction principle.
% ( since the standard induction tactic will not work?)

@@

After December, it remains to define the translation between GV and CP which I
envisage will be the bulk of the work, requiring several rules on typing
derivations; I plan to define these as inductive datatype constructors but I
shall take care to reuse any suitable libraries that are available, e.g. a
library for handling CPS translations. For this step of the work, I allocated
one month but this is a cautious estimate accounting some slack time for any
delays in preceding stages of the development; I suspect this stage will not
require a full month.

For the remainder, I shall look to extend the definitions of GV and CP using
the definitions I have provided. As discussed previously, it would be
interesting to study a system for supporting side-effecting functions (aka
closures) for session channels. In order to achieve this, operational
semantics for GV would need to be defined, and the recent draft by
\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
do this~\cite{Lindley:2014:SPS}. I allocated the remaining time (approx. two
months) to perform this extension because I believe it requires the most
development: extending existing definitions to handle side-effecting
functions, defining operational semantics, proving subject reduction and
related properties. Additionally, it will also require a theoretical treatment
prior to undertaking the development task, in contrast to the preceding
development work which mostly utilises results from the literature.

I shall review and update the above plan throughtout the project as I proceed
to the various milestones I have set out, and whenever I identify changing
circumstances.
